---
title: "Data processing"
author: "Anne Visser"
date: "2026-05-30"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(ellmer)
raw_data <- readRDS(here("data/raw_data.rds"))
data_complete <- readRDS(here("data/data_complete.rds"))
new_variables <- setdiff(colnames(data_complete), colnames(raw_data))
```

## How can chat history be saved for use in analysis?

In the prototype querychat app connected to the varInfo_synthetic table in the RVAT-database all the benchmark questions are clickable in the greeting, making sure that these exact questions are asked.

### Saving chat history
To benchmark the SQL-query generated by the LLM model there is a need for obtaining and storing the responds from the LLM model. Manually writing it down would be prone to errors or bias to correctly answered questions. To make storing the data easier a button was made within the querychat app to save the chat history. This button is a so called 'reactive value' within the Shiny app (https://shiny.posit.co/r/articles/build/reactivity-overview/), whenever an event happens to this object, a reaction will follow using the shiny app function observeEvent().  
In this case the reaction is getting the history of the current chat by accessing the Ellmer function get_turns() within the querychat app and turning this into a table where each row contains information of a complete cycle. A cycle starts with a user question (in this case a benchmark question) and ends with the LLM answer to the question. This logging function is written in *"prototype/0_logging_function.R"* which was written with the use of Mistral's LeChat. The prompt is shown in *"prototype/prompts/logging_function_prompt.txt"*.
The rows are then written or added to the file *"data/raw_data.rds"*. 

The get_turns() function is accessed through the shiny app server by utilizing the querychat reactive value 'client'. Notice the code below can only be used inside the querychat app server (step 3 in *"prototype/3_querychat_app_setup.R"*). `qcvals` contains the reactive values within querychat. 

``` {r names_do_not_include, eval=FALSE}
# Show reactive values within the querychat app.
names(qcvals)
## [1] "client" "sql"    "title"  "df" 

# The client is the chat object used (in this case the ellmer chat object)
class(qcvals$client)
## [1] "Chat" "R6" 

# Within this chat object there are several functions that can be used:
names(qcvals$client)
##  [1] "stream_async"          "stream"                "set_turns"            
##  [4] "set_tools"             "set_system_prompt"     "register_tools"       
##  [7] "register_tool"         "on_tool_result"        "on_tool_request"      
## [10] "last_turn"             "initialize"            "get_turns"            
## [13] "get_tools"             "get_tokens"            "get_system_prompt"    
## [16] "get_provider"          "get_model"             "get_cost"             
## [19] "clone"                 "chat_structured_async" "chat_structured"      
## [22] "chat_async"            "chat"                  "add_turn"             
## [25] ".__enclos_env__"   

```

The variables extracted from the chat history are shown below:  

**`r colnames(raw_data)[1]`**: The question asked by the user.  
**`r colnames(raw_data)[2]`**: The SQL-query generated by the LLM.  
**`r colnames(raw_data)[3]`**: The name of the tool used.  
**`r colnames(raw_data)[4]`**: If there is an error when running the tool the error is shown.  
**`r colnames(raw_data)[5]`**: The LLM model used.  
**`r colnames(raw_data)[6]`**: Total user/assistant turns (normally 4 turns: 1) The user prompt. 2) The LLM starting the tool call. 3) The toolcall information is returned to LLM. 4) The answer the LLM gives to the user). More turns are possible when the LLM didn't return a working query the first time.  
**`r colnames(raw_data)[7]`**: The response the LLM gives in the chatbox.  
**`r colnames(raw_data)[8]`**: The replicate number. This needs to be manually edited in the *"prototype/3_querychat_app_setup.R"* script.  
**`r colnames(raw_data)[9]`**: A timestamp to see what time and date the 'Save Chat History' button was clicked, to make it easier to check where the data fits in the timeline if any changes were made in the context for instance.   

### Enriching and cleaning raw data

In the script *"4_making_data_complete"* the raw data is cleaned and enriched. First the following variables were added:  

**`r new_variables[1]`**: Each question gets a number for easy filtering.  
**`r new_variables[2]`**: Chat cycles with timeout errors were not shown when clicking 'Save Chat History' button so had to be added manually. The fail_reason gives reason timeout for these chat cycles.  
**`r new_variables[3]`**: The question type is added per question.  
**`r new_variables[4]`**: The gold truth is the/a correct SQL-code, these were provided with the benchmark questions. These can be used to generate the golden truth output to compare to the output of the LLM query.  
**`r new_variables[5]`**: The gold truth without AS. In SQL 'AS' is used to name a column when aggregating data. In the gold_truth_no_AS variable, the gold truth is copied but the AS command is skipped. This is to prevent a low exact match score for the sole reason of query having a different generated column name. For instance when the gold_truth names a column where variants are counted var_count and the LLM names the column 'variant_count', but the rest of the query is the same, this will still give an exact match score when using gold_truth_no_AS to compare the LLM SQL-query to. 

After enriching, the script cleans the data.  It only keeps complete cycles (these are recognized by having a model name present) and cycles with benchmark questions (recognized by not having a question number). It also gets rid of any duplicate rows (exlcuding the timestamp) that are present when the 'Save Chat History' button is clicked more than once in a session. The table is then written to *"data/data_complete.rds"*, leaving the raw data file intact.

### Conclusion/Discussion
The logging is sufficient for scoring EM, EX and YES based on the final answer. It also gives some insight into how many turns an LLM needed to make a working query (total_turns), which may account for the difficulty of the benchmark question. 
The turns in between are not shown, so these cannot be analysed. This might mean certain vulnerabilities of the LLM are not noticed. It could happen that a model has a correct last answer, but tried several other answers with incorrect logic or hallucination. Since the built prototype is meant for research questions, it can't afford to use a model that frequently hallucinates or disregards instructions.
Furthermore the logging function should be evaluated (and maybe partially rewritten) by someone with knowledge of the Ellmer-package and R to check for any oversights in the function itself. 
Further inqueries can be done for the usefulness of the R-package vitals() for benchmarking.
